Asymmetric tail-association in extreme climatic events affects population-extinction risk

### 2019-10-14 MainText.Rmd
intro edited as of now, last para need to modify..Questions for DCR

<!--concept for correlation in the extreme tail: tail-association-->
Though earlier literature considered the influence of spatial and/ or temporal 
correlations in environmental variables on extinction risk of a metapopulation, to 
our knowledge the separate influence on extinction risk of "tail associations" between
environmental time series in different locations has hardly been considered. "Tail associations" are
the association between two random variables in their extreme values. If two positively correlated 
variables are more associated in their smaller values than larger, they are said to have stronger
"left-tail" or "lower-tail" association (Figure \ref{fig_intro_cop}A, B). Similarly, if two positively correlated variables are more associated in their larger values than smaller, they are 
said to have stronger "right-tail" or "upper-tail" association (Figure \ref{fig_intro_cop}C, D). 
Figure \ref{fig_intro_cop} shows two random variables can have a wide 
range of assciation from extreme to moderate degree in one of their 
extreme tails - often refers to having 'asymmetric tail' - but with a 
fixed Kendall's $\tau$. One can control the degree of "tail-asymmetry" - moderate 
to extreme - using "copula" (for details see @Ghosh_copula and \nameref{Methods}), a statistical 
tool generally used to model dependencies between correlated variables. 
This 'asymmetric tail' is commonly found in many 
bivariate or multivariate dependence structures for environmental, species-trait, 
phenological, population, community, and ecosystem functioning datasets 
and can help to interpret the causal mechanisms as well as 
to draw consequential inferences for such patterns [@Ghosh_copula]. For example, it 
is well known that spatial dependence of environmental
variables is an important driver of spatial dependence in fluctuating 
population time series (called 'spatial synchrony'). Therefore, one may expect 
the synchronous populations over a larger network would suppress for extreme bad 
years of limiting resources than moderately 'good' years. Any two such populations, that are 
affected badly in a similar manner due to common food shortage, are expected to be more
"left-tail" associated (Figure \ref{fig_intro_cop}A, B) rather than "right-tail" (Figure \ref{fig_intro_cop}C, D) as the environmental fluctuation 
time-series have more "left-tail" association for those extreme bad years
(i.e., 'Moran-effect' like mechanism). Similarly, during 
favorable years if pest metapopulations 
grow abundantly compared to other average years, population time series look like 
\ref{fig_intro_cop}C or D with stronger "right-tail" association, turning local 
outbreaks into regional epidemics. In a 
nutshell, "tail-asymmetry" could have profound effect in understanding 
risk for population viability or ecosystem function
considering their relation with extreme climatic events.

<!--How copula could be relevent in estimating extreme event's 
influence on extinction risk : introduce Taildep. concept-->
A recent study [@Ghosh_copula] used the above mentioned feature of "tail-asymmetry" 
to estimate extinction risk for a density-independent metapopulation model. The study 
showed "left-tail" associated density-independent metapopulations 
to be more prone to extinctions than "right-tail" associated one as we may expect species 
are more vulnerable to "catastrophs" (extreme bad years) than to "bonanzas" (extreme good years).
Present manuscript is an extension of that concept in estimating the extinction risk 
beyond the density-independent model. Population growth rate, when depending 
on its own density, serves as a more realistic model to regulate unbounded growth, hence 
models with "density-dependence" are often studied in population ecology to 
predict population viability over long-term period [@hassell1975density; @fowler1987review; 
@courchamp1999inverse]. Here, we will explore the general finding from @Ghosh_copula in the context of
the density-dependent metapopulation patches which experience same spatially correlated noise but having a variety of tail asymmetry (from extreme to moderate, Fig. \ref{fig_intro_cop}). It is expected that extreme left-tail associated noise might wipe out some metapopulation patches 
during unfavorable years unless they are rescued by the nearby populations. This scenario 
could be more complex and specific of the direction of association 
depending on the dynamics of the model (see \nameref{Results}).

<!--and even applicable for a simple age-structured population where its vital rates are influenced by ECEs. Following we will reintroduce the concept of copula in a manner that follows @Ghosh_copula.-->

<!--***DAN: One thing you probably want to include is the logic (see BIVAN) that if extreme events are getting stronger and more extreme/intense events are also more spatially widespread, then it seems reasonable to expect that spatial patterns of tail associations in environmental variables will also be changing. You may also want to mention (one sentence only) that we use the statistical concept of the copula to control tail dependence in our simulations – more will be said about copulas in the Methods.-->

@@

From ecological perspective, it is reasonable to expect that extreme cold winter during growing season
might reduce the adult reproduction rate as well as larval survival (i.e. stronger 
tail association between vital rates when temperature is too cold than normal) and 
thereby inflate the population viability.

Similarly, in context of 
density-dependent metapopulation models, if extreme environmental fluctuations across sites 
appear to be as synchronous (left tail dependent or right tail dependent between patches 
with same correlation as in Fig.....) then it could wipe out some metapopulation patches 
during unfavorable years unless they are rescued by the nearby populations. This scenario 
could be more complex and specific of the direction of association 
depending on the dynamics of the model (see \nameref{Results}).

Though tail-association 
property of a copula has been used for risk-analysis in other fields like finance, 
hydrology etc. [@alexander2008market; @li2013; @chen2011flood; @Goswami2018], but to our 
knowledge asymmetric tail association have never been considered 
in estimating the effect of ECEs through extinction risk of age structured 
model or density-dependent metapopulations.

>>>>

<!--concept for correlation in the extreme tail: tail-association-->
Though earlier literature considered the influence of spatial and/ or temporal 
correlations in environmental variables on extinction risk of a metapopulation, to 
our knowledge the separate influence on extinction risk of "tail associations" between
environmental time series in different locations has hardly been considered. "Tail associations" are
the association between two random variables in their extreme values. If two positively correlated 
variables are more associated in their smaller values than larger, they are said to have stronger
"left-tail" or "lower-tail" association (Figure \ref{fig_intro_cop}A, B). Similarly, if two positively correlated variables are more associated in their larger values than smaller, they are 
said to have stronger "right-tail" or "upper-tail" association (Figure \ref{fig_intro_cop}C, D). 
Figure \ref{fig_intro_cop} shows two random variables can have a wide 
range of assciation from extreme to moderate degree in one of their 
extreme tails - often refers to having 'asymmetric tail' - but with a 
fixed Kendall's $\tau$. One can control the degree of "tail-asymmetry" - moderate 
to extreme - using "copula", a statistical 
tool generally used to model dependencies between correlated variables (for details see @Ghosh_copula and \nameref{Methods}). 
This 'asymmetric tail' is commonly found in many 
bivariate or multivariate dependence structures for environmental, species-trait, 
phenological, population, community, and ecosystem functioning datasets 
and can help to interpret the causal mechanisms as well as 
to draw consequential inferences for such patterns [@Ghosh_copula]. For example, it 
is well known that spatial dependence of environmental
variables is an important driver of spatial dependence in fluctuating 
population time series (called 'spatial synchrony'). Therefore, one may expect 
the synchronous populations over a larger network would suppress for extreme bad 
years of limiting resources than moderately 'good' years. Any two such populations, that are 
affected badly in a similar manner due to common food shortage, are expected to be more
"left-tail" associated (Figure \ref{fig_intro_cop}A, B) rather than "right-tail" (Figure \ref{fig_intro_cop}C, D) as the environmental fluctuation 
time-series have more "left-tail" association for those extreme bad years
(i.e., 'Moran-effect' like mechanism). Similarly, during 
favorable years if pest metapopulations 
grow abundantly compared to other average years, population time series look like 
\ref{fig_intro_cop}C or D with stronger "right-tail" association, turning local 
outbreaks into regional epidemics. In a 
nutshell, "tail-asymmetry" could have profound effect in understanding 
risk for population viability or ecosystem function
considering their relation with extreme climatic events.

A recent study [@Ghosh_copula] used the above mentioned feature of "tail-asymmetry" 
to estimate extinction risk for a density-independent metapopulation model. The study 
showed "left-tail" associated density-independent metapopulations 
to be more prone to extinctions than "right-tail" associated one as we may expect species 
are more vulnerable to "catastrophs" (extreme bad years) than to "bonanzas" (extreme good years).
Present manuscript is an extension of that concept in estimating the extinction risk 
beyond the density-independent model. Population growth rate, when depending 
on its own density, serves as a more realistic model to regulate unbounded growth, hence 
models with "density-dependence" are often studied in population ecology to 
predict population viability over long-term period [@hassell1975density; @fowler1987review; 
@courchamp1999inverse]. Here, we will explore the general finding from @Ghosh_copula in the context of
the density-dependent metapopulation patches which experience environmental noises with same spatial correlation but with a variety of tail 
asymmetry (from extreme to moderate, Fig. \ref{fig_intro_cop}). It is expected 
that extreme left-tail associated noise might wipe out some metapopulation patches 
during unfavorable years unless they are rescued by the nearby populations. This scenario 
could be more complex and specific of the direction of association 
depending on the dynamics of the model (see \nameref{Results}).

Tail-asymmetry in estimating extinction risk can also be applied to an age structured model. From ecological perspective, it is reasonable to expect that extreme cold winter during growing season
might reduce the adult reproduction rate (fecundity) as well as larval survival. This scenario would resemble Fig.\ref{fig_intro_cop}(A, B) with stronger tail association between vital rates of two stages at their smaller values than larger ones and thereby inflate the population viability compared to a scenario as of Fig.\ref{fig_intro_cop}(C, D). <!--***Shyamolina: Should I tell this? 
It is reasonable to assume cold winter-temperature as the limiting factor 
for each age-structured population - thereby also showing "left-tail" asymmetry between 
the temperature and each age-population time series.(Ref: driver-synchrony relationship as shown in our aphid-plankton paper)--> Though tail-association 
property of a copula has been used for risk-analysis in other fields like finance, 
hydrology etc. [@alexander2008market; @li2013; @chen2011flood; @Goswami2018], but to our 
knowledge asymmetric tail association have never been considered 
in estimating the effect of ECEs with age structured 
model or density-dependent metapopulations.

@@
